---
title: "expectedCases: Projecting Cancer Incidence"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tutorial: expectedCases}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---



```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(expectedCases)
library(tibble)
library(ggplot2)
```


## Overview

The **expectedCases** package projects expected cancer incidence and mortality for a cohort using age-specific rates. This vignette shows how to (1) prepare a 5-year age-band cohort, (2) supply sex-specific incidence and mortality, (3) run projections, and (4) interpret/visualize results.

## Cohort requirements
Cohort counts must be in **5-year age bands** (e.g., 40–44, 45–49, …). Incidence may be in broader bands (e.g., 40–54, 55–69); the function maps those to single years internally. Open-ended bands (e.g., 85+) are supported and extended to `age_max`.


## Step 1. Create the baseline cohort


Each age band must represent a **5-year range** (e.g., “40–44”, “45–49”, etc.).
The function assumes equally spaced 5-year intervals for age expansion.  


```{r cohort, message=FALSE, warning=FALSE}
counts_5y <- tibble::tibble(
  age_band = c("40-44","45-49","50-54","55-59"),
  N        = c(2694, 3480, 4166, 4964)
)
counts_5y
```

---

## Step 2. Define incidence and mortality rates

Rates are specified separately for females and males.  
- **Incidence rates** are typically provided in *broader grouped age bands* (e.g., “0–24”, “25–39”, “40–54”, “55–69”, “70–85+”).  
- **Mortality rates**, by contrast, are usually specified in *detailed 5-year bands* (e.g., “40–44”, “45–49”, “50–54”, etc.).

Note that while the cohort (counts_5y) must always use 5-year age intervals (e.g., “40–44”, “45–49”),
the incidence tables may use broader age groups (e.g., “40–54”).
The function automatically splits these wider bands into matching 5-year intervals internally for calculation.

```{r rates, message=FALSE, warning=FALSE}
# Sex proportions
female_share <- 0.6
male_share   <- 0.4

# Incidence and mortality rates per 100k (simplified integers)
# IMPORTANT: Include an open-ended "85+" band so coverage reaches age_max (e.g., 100)

female_inc_wide <- tibble(
  band = c("40-54", "55-69", "70-84", "85+"),
  rate_per100k = c(39, 102, 278, 400)  # last value is illustrative; use your preferred rate
)

female_mort_5y <- tibble(
  band = c("40-44","45-49","50-54","55-59","60-64","65-69","70-74","75-79","80-84","85+"),
  rate_per100k = c(87, 138, 214, 309, 455, 707, 1164, 2104, 4133, 13700)
)

male_inc_wide <- tibble(
  band = c("40-54","55-69","70-84","85+"),
  rate_per100k = c(45, 148, 338, 500)  # last value is illustrative; use your preferred rate
)

male_mort_5y <- tibble(
  band = c("40-44","45-49","50-54","55-59","60-64","65-69","70-74","75-79","80-84","85+"),
  rate_per100k = c(155, 225, 337, 506, 789, 1190, 1888, 3239, 5974, 15415)
)
```

**Validation:** Age bands must be digits only (`NN-NN` or `NN+`). Typos like `4o-44` (letter **o**) are rejected with a clear error. Rates must cover **every** age in `age_min:age_max` exactly once (no gaps/overlaps).



## Step 3. Run the projection

## Function Arguments

| Argument | Type | Description |
|:--|:--|:--|
| `counts_5y` | `tibble` | Baseline cohort counts by **5-year age bands** (e.g., "40-44", "45-49"), with one column `age_band` and one column `N` (population size). |
| `female_share` | `numeric` | Proportion of the total cohort that is female (between 0 and 1). |
| `male_share` | `numeric` | Proportion of the total cohort that is male (between 0 and 1). Must satisfy `female_share + male_share = 1`. |
| `female_inc_wide` | `tibble` | Female **incidence rates** per 100,000 population, specified in broader age bands (e.g., "40-54", "55-69", "70-85"). The function maps these broader age bands to single-year values internally, assuming flat rates within each band. |
| `female_mort_5y` | `tibble` | Female **all-cause mortality rates** per 100,000 population, given in 5-year bands (e.g., "40-44", "45-49", …, "85+"). |
| `male_inc_wide` | `tibble` | Male incidence rates per 100,000 population, formatted like `female_inc_wide`. |
| `male_mort_5y` | `tibble` | Male all-cause mortality rates per 100,000 population, formatted like `female_mort_5y`. |
| `study_start` | `numeric` | Starting calendar year of the cohort (e.g., `2020`). |
| `study_end` | `numeric` | Ending year of data collection or baseline projection period (e.g., `2025`). Used to determine midpoint for internal timing. |
| `age_min` | `numeric` | Minimum age of the modeled cohort (default: `40`). |
| `age_max` | `numeric` | Maximum modeled age (default: `100`). Individuals exceeding this are dropped from the cohort. |
| `end_year` | `numeric` | Final projection year (e.g., `2040`). Determines length of simulation. |
| `begin_year` | `numeric` | Optional. Custom first projection year (default = midpoint of `[study_start, study_end]`). |
| `diag_years` | `numeric` vector | Optional. Specific years at which projections (e.g., 2022, 2030, 2040) are returned in the output. |
| `print_markdown` | `logical` | If `TRUE`, prints a formatted markdown summary table of expected cases in the console. Set `FALSE` for silent operation (recommended for vignettes). |



### Return Value

The function returns a **list** with two tibbles:

| Output | Description |
|:--|:--|
| `summary_tbl` | Aggregated expected cancer cases by sex and projection period. |
| `projection_tbl` | Detailed year-by-year projection showing `alive_start`, `new_cases_year`, `deaths_year`, `aged_out_year`, `alive_end`, and `cum_cases`. |


```{r run-projection, message=FALSE, warning=FALSE}
res <- run_cancer_projection(
  counts_5y,
  female_share = female_share, male_share = male_share,
  female_inc_wide, female_mort_5y,
  male_inc_wide,   male_mort_5y,
  study_start = 2020, study_end = 2025,
  age_min = 40, age_max = 100,
  end_year = 2040,
  diag_years = 2022:2040,
  print_markdown = FALSE
)

projection_tbl <- res$projection_tbl
summary_tbl    <- res$summary_tbl
```

Before interpreting results, it may help to understand the model structure used internally by `expectedCases`.

## Statistical Framework

The population projection in `expectedCases` follows a deterministic cohort-based model,  
where each age–sex group evolves over discrete yearly time steps.

For each age group \(a\), sex \(s\), and year \(t\):

\[
\begin{aligned}
\text{NewCases}_{a,s,t} &= N_{a,s,t} \times \lambda_{a,s,t}, \\
\text{Deaths}_{a,s,t}   &= N_{a,s,t} \times \mu_{a,s,t}, \\
\text{AliveEnd}_{a,s,t} &= N_{a,s,t} - \text{NewCases}_{a,s,t} - \text{Deaths}_{a,s,t}, \\
N_{a+1,s,t+1} &= \text{AliveEnd}_{a,s,t}.
\end{aligned}
\]

where:

| Symbol | Meaning |
|:--|:--|
| \(N_{a,s,t}\) | Individuals alive and cancer-free at the start of year \(t\). |
| \(\lambda_{a,s,t}\) | Age–sex–specific incidence rate (probability of developing cancer). |
| \(\mu_{a,s,t}\) | Age–sex–specific mortality rate (probability of death from any cause). |
| \(\text{NewCases}_{a,s,t}\) | Expected new cancer cases during year \(t\). |
| \(\text{Deaths}_{a,s,t}\) | Expected all-cause deaths during year \(t\). |
| \(\text{AliveEnd}_{a,s,t}\) | Number of individuals surviving cancer-free to the end of year \(t\). |




### Continuous-Time Analogy

In continuous time, the discrete recursion discretizes the continuous-time hazard formulation, which represents competing risks between cancer incidence and mortality.

\[
\frac{dN_{a,s}(t)}{dt} = -N_{a,s}(t) \times (\lambda_{a,s}(t) + \mu_{a,s}(t)).
\]

which represents competing risks between cancer incidence and mortality.  
In the discrete (annual) version used here, the yearly transition follows:

\[
N_{a+1,s,t+1} \approx N_{a,s,t} \times [1 - (\lambda_{a,s,t} + \mu_{a,s,t})].
\]



## Step 4. Interpret the results

The output includes:

* `summary_tbl`: summarized expected cases by sex and projection period
* `projection_tbl`: detailed year-by-year estimates (alive, new cases, deaths, cumulative cases)

```{r view-yearly, message=FALSE, warning=FALSE}
knitr::kable(summary_tbl, caption = "Expected cases by sex and period")

projection_tbl |>
  dplyr::select(Sex, year, alive_start, new_cases_year, deaths_year, aged_out_year, alive_end, cum_cases) |>
  dplyr::mutate(dplyr::across(where(is.double), ~ round(.x, 1))) |>
  knitr::kable(caption = "Year-by-year projection (rounded for display)")

```



```{r explanation, echo=FALSE,  message=FALSE, warning=FALSE}

  

knitr::kable(
  data.frame(
    Column = c("Sex", "year", "alive_start", "new_cases_year", 
               "deaths_year", "aged_out_year", "alive_end", "cum_cases"),
    Description = c(
      "Sex of the subgroup (Female or Male).",
      "Calendar year of projection.",
      "Number alive at the start of the year within modeled ages.",
      "Expected new cancer cases during that year.",
      "Expected all-cause deaths during that year.",
      "Individuals reaching `age_max` who exit the modeled range next year.",
      "Number of survivors remaining within modeled ages at year-end.",
      "Cumulative expected cancer cases up to and including that year."
    )
  ),
  caption = "Description of projection table columns in `projection_tbl`."
)
```



  
## Step 5. Visualize trends

We can plot the expected cumulative cases over time by sex.

```{r plot, fig.width=5, fig.height=3.2, dpi=200,  message=FALSE, warning=FALSE}
ggplot(projection_tbl, aes(year, cum_cases, color = Sex)) +
  geom_line(linewidth = 0.9) +
  labs(title = "Expected Cumulative Cancer Cases", x = "Year", y = "") +
  theme_minimal(base_size = 10) +
  theme(
    legend.position = "bottom",
    legend.title = element_blank(),
    panel.grid.minor = element_blank()
  )
```


## Help

* For help and documentation:
  
```r
?run_cancer_projection
```

## Session information for reproducibility

```r
sessionInfo()
```
